\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles

### 2023-03-18 Manuscript/manuscript.tex
Revisions

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[1,2]{Stephanie C. Hicks\thanks{Corresponding author}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA}
\affil[2]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}
\date{\today}

@@

%% INTRODUCTION ============================================================================================
\section{Introduction}
Visualization is an indispensable component of data analysis, providing clarity that connects quantitative evidence to key conclusions \cite{dagostinomcgowan_2022}. In biomedical research, visualization receives growing recognition as essential: many scientists rely on visualization to complete their cognitive process from analysis to insight, including analytic validation of automated pipelines and scientific communication \cite{odonoghue_2021}. However, an important challenge in biomedical resesearch is how to visualize increasingly complex, multi-dimensional data \cite{odonoghue_2010}.

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis (PCA) \fixme{add citation}, $t$-SNE \cite{hinton_2002} \fixme{spell out}, or UMAP \cite{becht_2019} \fixme{spell out}) and (ii) \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution \fixme{add citations} \cite{moffitt_2022}. These technologies are commonly referred to as single-cell/single-nucleus RNA-sequencing (sc/snRNA-seq) \fixme{add citations} and \textit{in situ} spatially-resolved transcriptomics (SRT) \fixme{add citations}.

A common and fundamental challenge with both of these visualizations is how to visualize multi-dimensional information in a 2D space. For example, in \textit{in situ} visualizations, we often want to create a spatial map to visualize a continuous (e.g. gene expression) or discrete (e.g. cell type or spatial domain) variable representing molecular information in the original spatial location. However, it is challenging to simultaneously visualize multi-dimensional data, such as information from disparate data domains (such as expression domain and spatial domain) or disparate data modalities (such as transcriptomics and proteomics) in the same plot. Currently, best practices for this include making two different plots displayed side-by-side (\textbf{Figure \textbf{\ref{fig:visual}A-B}}), one for gene expression and one for spatial domains. This creates cognitive gaps on how to associate the disparate information or how to interpret the biological findings of this multi-dimensional information regarding their (micro-)environment or colocalization. While interactive visualizations \cite{sriworarat_2023} \fixme{add more citations} or 3D visualizations \fixme{add citations} have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as printed work. Developing a static and unified visualization that enables the simultaneous display of multiple dimensions of information is crucial for biomedical research.

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/fig_grid.png}
\caption{\small \textbf{\texttt{escheR} improves spatial visualizations following the Gestalt principles}. (\textbf{A-B}) The traditional visualization displays the cortex layers (spatial domains) (\textbf{A}) and the colocalization status (\textbf{B}) in a side-by-side figure, creating challenges to connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \texttt{in-situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability. (\textbf{E}) \texttt{schex} creates overlapped convex hulls, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. (\textbf{F}) \texttt{escheR} plots hexagon-specific membership to improve interpretability.}
\label{fig:visual} 
\end{center}
\end{figure*}

@@

Gestalt principles\cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Developed in the 1920s by German psychologists Max Wertheimer, Kurt Koffka and Wolfgang Kohler, these principles helps humans perceive a set of individual elements as a whole.

Here, we leverage the principles to be able to visualize multi-dimensional data in a unified 2D plot. Our approach is to use the state-of-art data visualization framework \texttt{ggplot2} \cite{ggplot2} following the grammar of graphics \cite{wilkinson_2012} and map individual variables to different aesthetics to simultaneously display disparate variables. Specifically, we apply a figure-ground articulation to display two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. In practice, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map.

For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles (\textbf{Figure~\ref{fig:visual}C}), inspired by watercolor effect \cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground articulation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles \cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

Next, we give two use cases to exemplify some utility of the proposed spatial visualization: (i) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data \cite{huukimyers_2023}; (ii) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

\subsection{Multi-dimensional \textit{in situ} visualization}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex (DLPFC) \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), were identified to colocalize via the cell-cell communication analysis. In addition, data suggested Layer 6 was the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we applied \texttt{escheR} to create a multi-dimensional \textit{in situ} spatial map (\textbf{Figure \ref{fig:visual}A-B}) that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of genes \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are visualized in two side-by-side figures, our proposed visualization (\textbf{Figure \ref{fig:visual}D}) enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

@@

The application of the proposed framework is not limited to \textit{in situ} visualizations of spatially-resolved transcriptomics data. It is broadly applicable to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and $t$-SNE. To demonstrate, we applied the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, previous work proposed to apply hexagonal binning strategy to display the gene expression \cite{freytag_2020}. Here, the color-coded convex hulls are used to annotate different clusters of cells (\textbf{Figure~\ref{fig:visual}E}). However, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we replace the convex hulls with color-coded hexagons boundaries  (\textbf{Figure~\ref{fig:visual}F}) to avoid possible membership confusion. We note that our contribution to improve the visualization is easily implemented without any modification of \texttt{schex} as both are built upon the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

\fixme{add any abbreviations here}

@@

\begin{itemize}[nosep]
    \item Conceptualization: BG
    \item Data Curation: BG
    \item \fixme{add more here} use credit system (https://www.elsevier.com/authors/policies-and-guidelines/credit-author-statement)
\end{itemize}

@@

\subsection*{Availability of data and materials}
The spatial transcriptomics dataset was obtained from \texttt{spatialLIBD} (\href{http://research.libd.org/spatialLIBD}{research.libd.org/spatialLIBD}). The UMAP example follows the `using\_schex' vignette in the \texttt{schex} package (\href{https://www.bioconductor.org/packages/schex}{bioconductor.org/packages/schex}). The code that generates these figures is deposited at \href{https://github.com/boyiguo1/Manuscript_escheR}{github.com/boyiguo1/Manuscript\_escheR} \fixme{version XXX}.

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/fig_grid.png}
\caption{\small \textbf{\texttt{escheR} improves spatial visualizations following the Gestalt principles}. (\textbf{A-B}) The traditional visualization displays the cortex layers (spatial domains) (\textbf{A}) and the colocalization status (\textbf{B}) in a side-by-side figure, creating challenges to connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \texttt{in-situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability. (\textbf{E}) \texttt{schex} creates overlapped convex hulls, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. (\textbf{F}) \texttt{escheR} plots hexagon-specific membership to improve interpretability.}
\label{fig:embedding} 
\end{center}
\end{figure*}

>>>>

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[1,2]{Stephanie C. Hicks\thanks{$^*$Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA}
\affil[2]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}
\date{\today}

@@

%% INTRODUCTION ============================================================================================
\section{Introduction}
Visualization is an indispensable component of data analysis, providing clarity that connects quantitative evidence to key conclusions \cite{dagostinomcgowan_2022}. In biomedical research, visualization receives growing recognition as essential: many scientists rely on visualization to complete their cognitive process from analysis to insight, including analytic validation of automated pipelines and scientific communication \cite{odonoghue_2021}. However, an important challenge in biomedical research is how to visualize increasingly complex, multi-dimensional data \cite{odonoghue_2010}.

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis (PCA) \cite{hotelling_1933}, $t$-distributed Stochastic Neighbor Embedding ($t$-SNE) \cite{vandermaaten_2008}, or Uniform Manifold Approximation and Projection (UMAP) \cite{becht_2019}) and (ii) \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution \cite{kashima_2020, moffitt_2022}. These technologies are commonly referred to as single-cell/single-nucleus RNA-sequencing (sc/snRNA-seq) \fixme{add citations} and \textit{in situ} spatially-resolved transcriptomics (SRT) \fixme{add citations}.

A common and fundamental challenge with both of these visualizations is how to visualize multi-dimensional information in a 2D space. For example, in \textit{in situ} visualizations, we often want to create a spatial map to visualize a continuous (e.g. gene expression) or discrete (e.g. cell type or spatial domain) variable representing molecular information in the original spatial location. However, it is challenging to simultaneously visualize multi-dimensional data, such as information from disparate data domains (such as expression domain and spatial domain) or disparate data modalities (such as transcriptomics and proteomics) in the same plot. Currently, best practices for this include making two different plots displayed side-by-side (\fixme{\textbf{Figure \textbf{\ref{fig:visual}A-B}}}), one for gene expression and one for spatial domains. This creates cognitive gaps on how to associate the disparate information or how to interpret the biological findings of this multi-dimensional information regarding their (micro-)environment or colocalization. While interactive visualizations \cite{keller_2021, pardo_2022, sriworarat_2023} have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as printed work. Developing a static and unified visualization that enables the simultaneous display of multiple dimensions of information is crucial for biomedical research.

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/insitu.jpg}
\caption{\small \textbf{\texttt{escheR} improves spatial visualizations following the Gestalt principles}. (\textbf{A-B}) The traditional visualization displays the colocalization plot (\textbf{A}) and the cortex layers (spatial domains) plot (\textbf{B}) side-by-side, creating challenges to connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \texttt{in-situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability.}
\label{fig:visual} 
\end{center}
\end{figure*}

@@

Gestalt principles\cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Developed in the 1920s by German psychologists Max Wertheimer, Kurt Koffka and Wolfgang Kohler, these principles help humans perceive a set of individual elements as a whole.

Here, we leverage the principles to be able to visualize multi-dimensional data in a unified 2D plot. Our approach is to use the state-of-art data visualization framework \texttt{ggplot2} \cite{ggplot2} following the grammar of graphics \cite{wilkinson_2012} and map individual variables to different aesthetics to simultaneously display disparate variables. Specifically, we apply the figure-ground segmentation \cite{peterson_2010} in displaying two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. In practice, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map.

For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles (\fixme{\textbf{Figure~\ref{fig:visual}C}}), inspired by watercolor effect \cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground segmentation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles \cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

Next, we give two use cases to exemplify some utility of the proposed spatial visualization: (i) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex using spatial transcriptomics data \cite{huukimyers_2023}; (ii) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

\subsection{Multi-dimensional \textit{in situ} visualization}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), were identified to colocalize via the cell-cell communication analysis. In addition, data suggested Layer 6 was the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we applied \texttt{escheR} to create a multi-dimensional \textit{in situ} spatial map that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of genes \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are visualized in two side-by-side figures (\textbf{Figure \ref{fig:visual}A-B}), our proposed visualization (\textbf{Figure \ref{fig:visual}D}) enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

@@

The application of the proposed framework is not limited to \textit{in situ} visualizations of spatially-resolved transcriptomics data. It is broadly applicable to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and $t$-SNE. To demonstrate, we applied the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, previous work proposed to apply hexagonal binning strategy to display the gene expression \cite{freytag_2020}. Here, the color-coded convex hulls are used to annotate different clusters of cells (\textbf{Figure~\ref{fig:embedding}A}). However, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we replace the convex hulls with color-coded hexagons boundaries  (\textbf{Figure~\ref{fig:embedding}B}) to avoid possible membership confusion. We note that our contribution to improve the visualization is easily implemented without any modification of \texttt{schex} as both are built upon the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

\begin{itemize}[nosep]
    \item \textbf{PCA}: principal component analysis
    \item \textbf{$t$-SNE}: $t$-distributed Stochastic Neighbor Embedding
    \item \textbf{UMAP}: Uniform Manifold Approximation and Projection
    \item \textbf{EFNA5}: membrane-bound ligand ephrin A5
    \item \textbf{EPHA5}: ephrin type-A receptor 5
\end{itemize}

@@

\begin{itemize}[nosep]
    \item \textbf{Boyi Guo}:  Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing, Visualization
    \item \textbf{Stephanie C. Hicks}: Conceptualization, Resources, Writing - Review \& Editing, Visualization, Supervision, Project administration, Funding acquisition
\end{itemize}

@@

\subsection*{Availability of data and materials}
The spatial transcriptomics dataset was obtained from \texttt{spatialLIBD} (\href{http://research.libd.org/spatialLIBD}{research.libd.org/spatialLIBD}). The UMAP example follows the `using\_schex' vignette in the \texttt{schex} package (\href{https://www.bioconductor.org/packages/schex}{bioconductor.org/packages/schex}). The code that generates these figures is deposited at \href{https://github.com/boyiguo1/Manuscript_escheR}{github.com/boyiguo1/Manuscript\_escheR} (\textit{v1.0}).

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/embedding.jpg}
\caption{\small \textbf{\texttt{escheR} improves multi-dimensional embedding visualizations}. UMAP representation along $x$ and $y$ axis of \fixme{what data? just generally need to describe and cite it, if appli}. (\textbf{A}) The \texttt{schex} R/Bioconductor package uses color-coded convex hulls to annotate cell memberships, creating confusion when interpreting hexagons in overlapping hulls. (\textbf{B}) Our package \texttt{escheR} plots hexagon-specific membership to improve interpretability.}
\label{fig:embedding} 
\end{center}
\end{figure*}
